Joint Likelihood Deconvolution of Astronomical Images in the Presence of Poisson Noise

### 2023-11-30 src/tex/ms.tex
[showyourwork] overleaf sync

% Abstract with filler text
    \begin{abstract}
        We present a new method for Joint Likelihood Deconvolution (Jolideco) of a set of astronomical observations of the same sky region in the presence of Poisson noise. The method reconstructs a single flux image from a set of observations and given models for the point spread function (PSF) per observation, by optimizing the the posterior distribution based on th}
        joint Poisson likelihood of all observations under a patch based image prior. The patch prior is parameterised by a Gaussian Mixture model (GMM) which we trained on astronomical images with high signal to noise ratio, including data from the  James Webb Telescope as well as the GLEAM radio survey. During the reconstruction process the patch prior adapts to the patch structures in the data by finding the most likely GMM component for each patch in the image. By applying the method to simulated data we show that both the combination of multiple observations as well as the patch based prior lead to a much improved reconstruction quality in many different source scenarios as well as signal to noise regimes. We show also that the method yields superior reconstruction quality to alternative standard methods such as the Richardson-Lucy method. We also show the results of the method applied to example data from the \chandra observatory as well as the \ferminospace. By comparing the measured width of a counts based and the corresponding Jolideco flux profile we find the deconvolved width to be consistent with the theoretical expectation derived from the known width of the PSF.
    \end{abstract}

@@

    \section{Implementation}
    \subsection{Jolideco Framework}
    The goal of the reconstruction process is to optimize the a-posteriori likelihood defined by Equation~\ref{eq:total}. Given that there is a parameter for each pixel in the image,  
    this is a high dimensional optimization problem. Differentiable programming frameworks such as \texttt{PyTorch}~\citep{Pytorch2019} allow for solving high dimensional optimization problems using back-propagation and adapted optimization methods such as stochastic gradient decent (SGD). The latter is also applicable in our case, as we work with multiple observations at the same time, where a subset of the observations represents a \enquote{mini batch} in terms of ML.

@@

        \item[\bf M2.] \textbf{Jolideco (Uniform, n=10)}: we use \jolideco with a uniform prior \dvd{on $\mathbf{x}$}, \dvd{running the optimizer for} 10 iterations. We consider this as a baseline method reference similar to RL. While the optimization procedure is different, we expect similar behaviour for the local correlations of pixels. This ad hoc approach of early stopping is often used in practice to regularise the RL output.

        \item[\bf M3.] \textbf{Jolideco (Uniform, n=1000)}:  we use \jolideco with a uniform prior \dvd{on $\mathbf{x}$}, but running \dvd{the optimzier} for 1000 iterations. This is the same as above but without early stopping. This gives rise to the issue of RL decomposing extended emission into point sources.

        \item[\bf M4.] \textbf{Jolideco (Patch, GLEAM v0.1)}: we use \jolideco with a patch prior learned on the GLEAM data, described in Section~\ref{sssec:gleam-radio-data} and the patch normalisation described by Equation~\ref{eq:patch-norm}. We use cycle spinning and 2000 \enquote{training epochs}. Where one training epoch corresponds to a full iteration over all observations. By visualizing the value of the log-posterior we verified

@@

        \item[\bf M5.] \textbf{Jolideco (Patch, Zoran-Weiss)}: we use \jolideco with the patch prior of \cite{Zoran2011} and the mean subtracted patch norm \dvd{is this the same as ``the patch normalisation described by Equation~\ref{eq:patch-norm}''? Best to use the same language for clarity.}. This is for comparison to the original \dvd{be more specific} method and the method used in \cite{Bouman2016}. We use a maximum image norm with the maximum values defined by the ground truth. \dvd{Is this the convergence criterion? Do you describe what this means somewhere?}\footnote{This is not completely fair, as it requires knowledge of the brightest expected pixel in $\mathbf{x}$. In practice this could be estimated from the brightest pixel in the data and the shape of the PSF.}
    \end{enumerate}

>>>>

% Abstract with filler text
    \begin{abstract}
        We present a new method for Joint Likelihood Deconvolution (\jolideco) of a set of astronomical observations of the same sky region in the presence of Poisson noise. The method reconstructs a single flux image from a set of observations and given models for the point spread function (PSF) per observation, by optimizing the the posterior distribution based on the joint Poisson likelihood of all observations under a patch based image prior. The patch prior is parameterised by a Gaussian Mixture model (GMM) which we trained on astronomical images with high signal to noise ratio, including data from the  James Webb Telescope as well as the GLEAM radio survey. During the reconstruction process the patch prior adapts to the patch structures in the data by finding the most likely GMM component for each patch in the image. By applying the method to simulated data we show that both the combination of multiple observations as well as the patch based prior lead to a much improved reconstruction quality in many different source scenarios as well as signal to noise regimes. We show also that the method yields superior reconstruction quality to alternative standard methods such as the Richardson-Lucy method. We also show the results of the method applied to example data from the \chandra observatory as well as the \ferminospace. By comparing the measured width of a counts based and the corresponding \jolideco flux profile we find the deconvolved width to be consistent with the theoretical expectation derived from the known width of the PSF.
    \end{abstract}

@@

    \section{Implementation}
    \subsection{\jolideco Framework}
    The goal of the reconstruction process is to optimize the a-posteriori likelihood defined by Equation~\ref{eq:total}. Given that there is a parameter for each pixel in the image,  
    this is a high dimensional optimization problem. Differentiable programming frameworks such as \texttt{PyTorch}~\citep{Pytorch2019} allow for solving high dimensional optimization problems using back-propagation and adapted optimization methods such as stochastic gradient decent (SGD). The latter is also applicable in our case, as we work with multiple observations at the same time, where a subset of the observations represents a \enquote{mini batch} in terms of ML.

@@

        \item[\bf M2.] \textbf{\jolideco (Uniform, n=10)}: we use \jolideco with a uniform prior \dvd{on $\mathbf{x}$}, \dvd{running the optimizer for} 10 iterations. We consider this as a baseline method reference similar to RL. While the optimization procedure is different, we expect similar behaviour for the local correlations of pixels. This ad hoc approach of early stopping is often used in practice to regularise the RL output.

        \item[\bf M3.] \textbf{\jolideco (Uniform, n=1000)}:  we use \jolideco with a uniform prior \dvd{on $\mathbf{x}$}, but running \dvd{the optimzier} for 1000 iterations. This is the same as above but without early stopping. This gives rise to the issue of RL decomposing extended emission into point sources.

        \item[\bf M4.] \textbf{\jolideco (Patch, GLEAM v0.1)}: we use \jolideco with a patch prior learned on the GLEAM data, described in Section~\ref{sssec:gleam-radio-data} and the patch normalisation described by Equation~\ref{eq:patch-norm}. We use cycle spinning and 2000 \enquote{training epochs}. Where one training epoch corresponds to a full iteration over all observations. By visualizing the value of the log-posterior we verified

@@

        \item[\bf M5.] \textbf{\jolideco (Patch, Zoran-Weiss)}: we use \jolideco with the patch prior of \cite{Zoran2011} and the mean subtracted patch norm \dvd{is this the same as ``the patch normalisation described by Equation~\ref{eq:patch-norm}''? Best to use the same language for clarity.}. This is for comparison to the original \dvd{be more specific} method and the method used in \cite{Bouman2016}. We use a maximum image norm with the maximum values defined by the ground truth. \dvd{Is this the convergence criterion? Do you describe what this means somewhere?}\footnote{This is not completely fair, as it requires knowledge of the brightest expected pixel in $\mathbf{x}$. In practice this could be estimated from the brightest pixel in the data and the shape of the PSF.}
    \end{enumerate}
